Towards a standardized evaluation of imputation methodology

### 2022-09-20 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
Finish all but one TODO in reviewer responses

% In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion), imputation allows analysts to account for the sources of uncertainty that are related to the missingness problem. With the popular variant multiple imputation \citep{rubi76}, each missing value is replaced by a plausible data value several times, resulting in multiple completed data sets. The estimated parameters of interest for each of these completed sets can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. [TODO: add that if the goal is statistical inference, the uncertainty due to missingness should be quantified/captured/propagated by the missing data method. Refer to \citet{sper20} for the context of prediction, where multiple imputation is not always necessary.] [TODO: paraphrase ``the quality of a technique can only be evaluated with respect to the aims of the problem/s it's intended to solve in practice".]

@@

% Especially with novel imputation methods being propagated from the fields of machine learning and artificial intelligence, the differences may become more pronounced. Although these promising new methods seem to yield even sharper imputations than now-standard imputation methods, the comparison may not be fair due to the simulation setup.

The purpose of this paper is threefold: First, we raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as reviewers for statistical journals. Second, we provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground, but is in no way intended as an absolute solution. This identifies the third purpose of our paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties. 
% [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

% [TODO: split up into design, execution and reporting? \citep{pawe22}]

@@

% [TODO: describe problems with unclear simulation parameters, extrapolation beyond scope of simulations, QRPs \citep{pawe22} and misinterpretation \citep[see "investigator bias"][]{gree17}.]

% Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

% The comparative truth of the estimand(s) is determined by the choice of data-generating mechanism and whether sampling variance is included in the simulation setup, see Table \ref{table:dgm_truth}.

% If sampling variance cannot be omitted from the simulation scheme, multiple samples should be drawn from the data-generating mechanism (i.e., one sample per simulation repetition, a standard simulation setup).

% If sampling variance is not of interest, a single complete data set can be obtained from the data-generating mechanism (i.e., one sample for all simulation repetitions).

% \begin{table}[tb]
% \begin{center}
% \caption{Source of the comparative truth in simulation studies.}
% \label{table:dgm_truth}
% \begin{tabular}{lll}
% \hline
%                & With sampling variance      & Without sampling variance \\
%                & (multiple samples drawn)    & (one sample drawn) \\
% \hline  
% Model-based simulation   & data-generating model         & single sample \\
% Design-based simulation  & sufficiently large data set   & single sample \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}
%

@@

% [TODO: split up into data and missingness part of the DGM, with choices between data sources, missingness mechanisms and proportions. And divide problems between simulation and interpretation, see Reviewer 2: "whether method A (e.g. imputation) outperforms method B (e.g. listwise deletion) depends on the estimand, purpose of the study, used distributions (in the DGP, imputation model, possibly also analysis model), and simulation setup (e.g. dependency structure) - if a human misinterprets findings in a comparison of method A and B, then this is a cognitive problem and not the problem on how missingness was generated in the DGP."] 
%
%[TODO; describe problems with data generation and interpretation: If the source of the data is model-based, the correlations in the data may be so high that there is no effective missingness due to the predictive power in the other variables, and the data source may be unfair if the data generating model is equal to one of the imputation models under evaluation. With a design-based simulation, the data source is unlikely to have no missingness and the choice of initial missing data method may propagate into the evaluations.]

@@

% 1a. choose DGM (if model-based: choose n_obs, n_var, relations between var, var types, etc; if design-based: choose n_obs, missing data handling strategy)
% 1b. decide if sampling variance is needed
% 1c. draw complete sample(s)

@@

% [TODO: Make clear in this section that missingness mechanisms can be spurious (spurious MAR when correlations are too low, or when missingness proportion is too low). In reporting, the functional form or type of missingness is often omitted. As advice, MCAR can be a very informative mechnism for reference and MNAR may be useful for practice. Missingness proportions can be misinterpreted. Missingness pattern may have unexpected consequences (e.g., univariately generated multivariate missingness). In interpretation, watch out for generalizations from unintended special cases.]

% Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data-generating mechanism, some form of missingness therefore has to be induced.

@@

% \begin{table}[tb]
% \begin{center}
% \caption{Missingness mechanisms. [TODO: add refs \citet{seam13, meal15, dore18} for clarification and alternatives outlined in \citet{moha21, scho21, more18}]}
% \label{table:mech}
% \begin{tabular}{lll}
% \hline
%       & Interpretation and consequences of the mechanism \\
% \hline  
% MCAR  & Missing completely at random: The probability to be missing is the same for all \\
%       & cases. In other words, the missing values are missing at random and the observed \\
%       & values are observed at random. \\ \noalign{\smallskip}
% MAR   & Missing at random: The probability to be missing is the same within groups of \\
%       & cases defined by the observed data only. The essence is that observed relations in \\
%       & the data inform the missingness, such that MCAR may be assumed within the \\ 
%       & observed groups. \\ \noalign{\smallskip}
% v-MAR & Conditional on the fully observed variables missingness occurs at random 
%       & \citep[][p. 1026]{moha21}.\\
% MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
%       & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
%       & inferred from the observed data. This is considered non-ignorable missingness \\
%       & \citep[see e.g.][]{rubi76}. \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}

@@

% Moreover, the value of the actual missingness proportion may be diffusing too. Some studies use only 10\% missingness whereas other studies push the limits to additionally investigate the performance under missingness proportions of at least 50\%. The inconsistent display of simulation conditions may impact the objectivity of meta-evaluations over imputation methods, as one method's performance may appear to be favorable because of less stringent simulation conditions. This ultimately may lead to statisticians recommending a less efficient method to applied researchers, thereby limiting the efficiency of the imputation approach and unnecessarily lowering the statistical power.

% Another aspect of missingness patterns that may misguide evaluations is the complexity of the patterns across variables. It is hardly reasonable to imagine empirical data with only one incomplete variable, yet some simulation studies rely on univariate missingness patterns anyways.

@@

% Missingness mechanisms are usually assumed to be random (MAR; see Table \ref{table:mech} for definitions). However, the missingness mechanisms that are induced in simulation studies do not necessarily match any true missingness generating mechanisms in incomplete empirical data, nor the mechanisms that are typically assumed when imputing such data. On the one hand, there is MCAR missingness, which is unlikely to be the true missingness generating mechanism, but would yield insightful evaluations in a simulation setting. On the other hand, there is MNAR, which is arguably a quite reasonable source of many real-world missingness scenarios but is generally ignored except for those evaluations that are specifically targeted at non-ignorable applications. A disconnect between induced missingness, real missingness, and assumed missingness may result in simulation studies that are not as informative as they could be.

@@

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to mimic a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the true data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is generated randomly (see Figure 1). In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

% 2a. choose missingness mechanism(s): at least consider MCAR, vary MAR types
% 2b. choose missing data pattern(s): introduce non-monotone multivariate missingness
% 2c. choose missingness proportion(s): at least consider 10, 25 and 50\%

% First, one should always consider MCAR missingness (see Table \ref{table:mech} for definitions). This mechanism should be used as a reference condition. Every imputation method must be able to yield valid imputations under MCAR, both in terms of distributional characteristics as well as statistical inference.
% 
% Next, missing data should be induced conform a model that is dependent on the observed data (i.e. a MAR mechanism). Arguably, MAR is the most often-assumed mechanism in practice, and should thus be evaluated carefully.

@@

% \begin{table}[htb]
% \begin{center}
% \caption{Suggested univariate missingness proportions. [TODO: change this to 'more than one missingness proportion, depending on the context' and/or add references for the given proportions (e.g. about  that 30\% is a good range for MI).]}
% \label{table:prop}
% \begin{tabular}{lll}
% \hline
%       & Reasoning for suggested missingness proportion \\
% \hline  
% 10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
%       & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
%       & the true data inference. Performance of a missing data method should at least be \\
%       & acceptable for most missing data problems. \\ \noalign{\smallskip} % & \\
% 25\%  & This is a fair amount of missingness and will, depending on the observed data \\
%       & information, have a noticeable influence on the completed data inference. When \\
%       & compared to the condition with 10\% missingness, the inference obtained under \\
%       & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
%       & should increase), but estimates should still be properly covered and the statistical \\
%       & properties of the missing data method should be sound. In practice, at least to our \\
%       & at least to our experience in social sciences and official statistics, 25\% \\
%       & univariate missingness can easily be considered as a realistic missingness percentage. \\ \noalign{\smallskip} % & \\
% 50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
%       & severely. Depending on how the missingness mechanism interacts with the simulated \\
%       & data, some imputation techniques may yield estimates that are under-covered such that \\
%       & the completed data inference should not be deemed valid anymore. If a method yields \\
%       & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
%       & properties of the imputation methodology are sound. \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}

@@

% \subsection{Imputation methods}

% 3a. Define missing data methods: at least consider complete case analysis as benchmark
% 3b. Choose imputation model parameters: appropriate n_imp and n_it (if applicable), maybe vary predictor matrices?
% 3c. Impute the data
% 3d. Obtain estimates

\subsection{Performance evaluation}
The evaluation criteria used to assess imputation performance vary from one simulator to another. This is not surprising as people from different fields could have a different focus on the problem at hand. There are, however, some overarching issues when assessing imputation method performance. In the first place, there are pitfalls in the evaluation of the estimates that are obtained by fitting the analysis model after imputation. For example, focusing on one performance measure over another, may yield different conclusions about imputation efficacy. In the second place, diagnostic evaluation of the generated imputations is often left out of simulation study results. Identifying problems with the imputation-generating process (e.g., an iterative imputation algorithm) may offer explanations for under-performance of imputation methods. Such evaluations, however, are typically omitted and valuable insights into the imputation method(s) may then be overlooked. %(i.e. the complete-data model used to estimate the comparative truth). % The choice of performance measures might inadvertently influence simulation results.

@@

(i) The estimates should preferably be unbiased. Note that the way bias is calculated should be carefully chosen and described, since this can greatly influence the interpretation of the results. A negligible absolute bias, for example, may already yield an almost infinite relative bias if the true value of the estimand is zero. In most cases, unbiased estimates may be expected under a MCAR missingness mechanism which can be easily verified by including MCAR missingness to the scope of missingness investigations.

@@

% "Note that Neyman's original description of confidence intervals defined the property of randomisation validity as exactly 100(1_alpha) of intervals containing theta.44-46 Confidence validity is the property that the true percentage is at least 100(1_alpha). This latter definition is less well known than the former, with the result that overcoverage and undercoverage are sometimes regarded as similarly bad.47 Of course, randomisation validity would usually be preferred over confidence validity because it implies shorter intervals, but this is not always the case. There are examples of procedures that return both shorter intervals and higher coverage.45, 46" ADEMP paper

@@

(iv) Predictive accuracy may be quantified using the root mean squared error (RMSE). We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values and it may increase the rate of false positives \citep[Chapter 2.6][]{buur18}. However, if a study is aimed at obtaining predictions, and inferential validity is not of interest, the RMSE of the predicted values can be compared between method, which may yield valuable information about the methods' efficiency in terms of both accuracy and precision.

@@

% 4a. Evaluate imputations: check failed methods, algorithmic convergence (if applicable)
% 4b. Check imputation model fit: anomalies in distributional characteristics, plausibility of imputed values (if requested), PPC

% 4c. Apply performance measures: at least consider bias, coverage, CI length, if applicable also RMSE (at prediction and/or cell level)
% fraction of missing information (= model parameter), variance inflation

@@

% It could be convincing to demonstrate a method's applicability to real-world missing data problems. This can, for example, be achieved by obtaining and imputing a fully observed set of variables, wherein the missingness is mimicked from a similar, incomplete set. Alternatively, an incomplete set could be obtained and truth could be established by removing the incomplete cases from the data. However, the real-world missingness would then be omitted.

% \subsection{Reporting}

% 
% \begin{table}[tb]
% \begin{center}
% \caption{Steps to consider in imputation simulation studies.}
% \label{table:steps}
% \begin{tabular}{lll}
% \hline
%                         & Simulation aspects under consideration \\
% \hline  
% 1. Set scope            & Aim(s), missing data method(s) to evaluate, simulation design \\
% 2. Obtain truth         & Data-generating mechanism(s), estimand(s), sampling variance \\
% 3. Induce missingness   & Missingness mechanism(s), missing data pattern(s) \\
% 4. Apply methods        & Imputation method(s), analytic method(s) \\
% 5. Evaluate imputations & Imputation-generating process, imputation model fit \\
% 6. Evaluate performance & Performance measures, qualitative judgment \\
% 7. Report               & Text, visualization, checklist, simulation script \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}

>>>>

@@

The purpose of this paper is threefold: First, we raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as reviewers for statistical journals. Second, we provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground, but is in no way intended as an absolute solution. This identifies the third purpose of our paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties.

@@

@@

@@

@@

@@

@@

@@

@@

@@

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to mimic a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the true data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is generated randomly (see Figure 1). In fact, this would amount to one of the special cases described by \citet{buur18} under which complete case analysis would be more efficient than imputation: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

@@

@@

\subsection{Performance evaluation}
The evaluation criteria used to assess imputation performance vary from one simulator to another. This is not surprising as people from different fields could have a different focus on the problem at hand. There are, however, some overarching issues when assessing imputation method performance. In the first place, there are pitfalls in the evaluation of the estimates that are obtained by fitting the analysis model after imputation. For example, focusing on one performance measure over another, may yield different conclusions about imputation efficacy. In the second place, diagnostic evaluation of the generated imputations is often left out of simulation study results. Identifying problems with the imputation-generating process (e.g., an iterative imputation algorithm) may offer explanations for under-performance of imputation methods. Such evaluations, however, are typically omitted and valuable insights into the imputation method(s) may then be overlooked.

@@

(i) The methods should preferably be unbiased. Note that the way bias is calculated should be carefully chosen and described, since this can greatly influence the interpretation of the results. A negligible absolute bias, for example, may already yield an almost infinite relative bias if the true value of the estimand is zero. In most cases, unbiased estimates may be expected under a MCAR missingness mechanism which can be easily verified by including MCAR missingness to the scope of missingness investigations.

@@

@@

(iv) Predictive accuracy may be quantified using the root mean squared error (RMSE). We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values and it may increase the type I error rate of the statistical inferences \citep[Chapter 2.6][]{buur18}. However, if a study is aimed at obtaining predictions, and inferential validity is not of interest, the RMSE of the predicted values can be compared between method, which may yield valuable information about the methods' efficiency in terms of both accuracy and precision.

@@

@@
